---
title: "Example"
author: "Jon Ahlinder"
date: "2024-10-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(dplyr)
require(ggplot2)
library(forcats)
library(brms) # for analysis of PCs
library(nadiv) # for NRM calculation
```

# Example analysis of a Scots pine field trial


## Read data

```{r}
######### adjusted phenotypic records
dataF264_v2 <- 
  read.table(file = "../data/data_F264.txt",sep=",",header = T)
######### pedigree
pedigreeF264 <- read.table(file = "../data/pedigree_F264.txt",sep=",",header=T)
```

## Select a subset of data for example analysis
Note that you can impute missing data instead of just removing genotypes. The data is normalized as well.
```{r}
dataF264_v2 <- dataF264_v2 %>%
  na.omit() %>%
  mutate_each_(list(~scale(.) %>% as.vector),
                                  vars = 3:ncol(dataF264_v2))
```
## PCA

```{r}
results <- prcomp(dataF264_v2[,3:ncol(dataF264_v2)])
results$Genotype_id <- dataF264_v2$Genotype_id
# Access the proportion of variance explained by each PC
var_explained <- results$sdev^2 / sum(results$sdev^2)
# Create a data frame for plotting
plot_data <- data.frame(
  PC = as.factor(paste0("PC",1:length(var_explained))),
  VarianceExplained = var_explained*100
)
# Reorder the factor levels of PC based on VarianceExplained
plot_data <- plot_data %>%
  mutate(PC = fct_reorder(PC, VarianceExplained, .desc = TRUE))

# Create a bar plot using ggplot2
ggplot(plot_data, aes(x = PC, y = VarianceExplained)) +
  geom_bar(stat = "identity", fill = "steelblue",color = "black") +
  labs(
    x = "",
    y = "Variance explained (%)",
    title = ""
  ) +
  theme_minimal() + 
  theme(axis.text = element_text(size=11, color = "black"), axis.title = element_text(size = 11),
        axis.text.x = element_text(angle = 45))
# loading plot
# PC1 vs PC2
PC <- as.data.frame(results$rotation)
# fix labels
x_label <- sprintf("PC1 (%.1f%%)", plot_data$VarianceExplained[1])
y_label <- sprintf("PC2 (%.1f%%)", plot_data$VarianceExplained[2])
ggplot(PC, aes(x = PC1, y = PC2)) +
  geom_point(size=2) +
   theme(axis.text = element_text(size = 11,color = "black"),legend.text = element_text(size = 11)) +
  theme_minimal() +
  geom_text(
    label = rownames(PC),
    nudge_x=0.03, nudge_y=0.03,
    check_overlap=T,
color = "black",
  size = 3)  +
  labs(color='Trait category') +
  xlab(x_label) +
  ylab(y_label)
```

## Calculate numerator relationship matrix

```{r}
dataF264_v2 <- merge(dataF264_v2,pedigreeF264,by = "Genotype_id")
ord <- as.data.frame(results$x) %>%
  mutate(Genotype_id = results$Genotype_id)
dataF264_v2 <- merge(dataF264_v2,ord,by = "Genotype_id")
```

## Analyse PC components using Rstan via BMRS package
We use bmrs which runs Bayesian Multilevel Models using Stan. NB I only use the first three principal components here, but for optimal performance, all should be analysed. 
```{r}
ped_tmp <- data.frame(Genotype_id = dataF264_v2$Genotype_id, sire = dataF264_v2$Dad_id, dam = dataF264_v2$Mum_id) %>% 
  prepPed()
Amat <- as.matrix(makeA(ped_tmp)) # Create A matrix

bf_PC1 <- bf(PC1 ~ 1 + (1 | a | gr(Genotype_id, cov = Amat)))
bf_PC2 <- bf(PC2 ~ 1 + (1 | a | gr(Genotype_id, cov = Amat)))
bf_PC3 <- bf(PC3 ~ 1 + (1 | a | gr(Genotype_id, cov = Amat)))

############## PC1
brms_m1.pc1 <- brm(
  bf_PC1, 
  data = dataF264_v2,
  data2 = list(Amat = Amat),
  family = gaussian(),
  chains = 6, 
  cores = 6
)
plot(brms_m1.pc1)
summary(brms_m1.pc1)
mcmc_plot(brms_m1.pc1, type = "acf")
############## PC2
brms_m1.pc2 <- brm(
  bf_PC2, 
  data = dataF264_v2,
  data2 = list(Amat = Amat),
  family = gaussian(),
  chains = 6, 
  cores = 6
)
plot(brms_m1.pc2)
summary(brms_m1.pc2)
mcmc_plot(brms_m1.pc2, type = "acf")
############## PC3
brms_m1.pc3 <- brm(
  bf_PC3, 
  data = dataF264_v2,
  data2 = list(Amat = Amat),
  family = gaussian(),
  chains = 6, 
  cores = 6
)
plot(brms_m1.pc3)
summary(brms_m1.pc3)
mcmc_plot(brms_m1.pc3, type = "acf")

###################### heritability ###############################
v_individual <- (VarCorr(brms_m1.pc1, summary = FALSE)$Genotype_id$sd)^2
v_r <- (VarCorr(brms_m1.pc1, summary = FALSE)$residual$sd)^2
h.pc1 <- as.mcmc(v_individual / (v_individual + v_r))
summary(h.pc1)
v_individual <- (VarCorr(brms_m1.pc2, summary = FALSE)$Genotype_id$sd)^2
v_r <- (VarCorr(brms_m1.pc2, summary = FALSE)$residual$sd)^2
h.pc2 <- as.mcmc(v_individual / (v_individual + v_r))
summary(h.pc2)
v_individual <- (VarCorr(brms_m1.pc3, summary = FALSE)$Genotype_id$sd)^2
v_r <- (VarCorr(brms_m1.pc3, summary = FALSE)$residual$sd)^2
h.pc3 <- as.mcmc(v_individual / (v_individual + v_r))
summary(h.pc3)

plot(h.pc1)
plot(h.pc2)
plot(h.pc3)

```


## Obtain EBV on original phenotypic scale
```{r}
############################## Backtransform
### df with EBVs for each tree
G_ebv
data.pc.b <- G_ebv %>%
  select(c(PC1,PC2,PC3)) %>%
  as.matrix()
rotation <- as.matrix(results$rotation)
mu <- colMeans(dataF264_norm[,3:17])
EBV.pc <- as.data.frame(data.pc.b %*% t(rotation) + mu)
EBV.pc$Genotype_id <- selindex$Individual

```